---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geomultistar

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/josesamos/geomultistar.svg?branch=master)](https://travis-ci.com/josesamos/geomultistar)
<!-- badges: end -->

The goal of `geomultistar` is to build queries with geographical data in `multistar` format, obtained using the `starschemar` package. 

## Installation

You can install the released version of geomultistar from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("geomultistar")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("josesamos/geomultistar")
```

Multidimensional Queries Enriched with Geographic Data

Multidimensional systems allow complex queries to be 
    carried out in an easy way. The geographical dimension, 
    together with the temporal dimension, plays a fundamental 
    role in multidimensional systems. Through this package, vector 
    layers can be associated to the attributes of geographic 
    dimensions, so that the results of multidimensional queries 
    can be obtained directly as vector layers. The multidimensional 
    structures on which we can define the queries can be created 
    from a flat table with the starschemar package or imported 
    directly using functions from this package.


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(geomultistar)
## basic example code
```


